---
title: "Blog Post 4 - EDA and Modeling Section 1"
subtitle: ""
author: "Group 1"
date: "2024-11-06"
date-modified: "2024-11-06"
draft: false
description: ""
execute:
  enabled: false
---


The road with the highest number of crashes is unnamed, with just under 125 crashes recorded. The road with the second-highest number of crashes is Georgia Avenue, which has significantly fewer crashes, totaling just under 50

```{r p1}
# The top 10 roads with the highest number of crashes 
library(dplyr)
library(ggplot2)

cleaned_dataset <- read.csv("dataset/cleaned_dataset.rds")

road_crash_data <- cleaned_dataset %>%
  filter(!is.na(`Road.Name`)) %>%
  count(`Road.Name`, sort = TRUE) %>%
  top_n(10, n)

ggplot(road_crash_data, aes(x = reorder(`Road.Name`, n), y = n)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 10 Roads with the Most Crashes",x = "Road Name", y = "Number of Crashes") +
  theme_minimal()


```



```{r p2}
# Distribution of Simplified Surface conditions by collision type

library(dplyr)
library(ggplot2)

simplified_barplot_data <- cleaned_dataset %>%
  mutate(Simplified_Surface = case_when(
    Surface.Condition %in% c("Dry", "DRY") ~ "Dry",
    Surface.Condition %in% c("Wet", "WET", "Water (standing, moving)", "WATER(STANDING/MOVING)") ~ "Wet",
    Surface.Condition %in% c("Ice", "ICE", "Ice/Frost", "Snow", "SNOW", "Slush", "SLUSH") ~ "Snow/Ice",
    Surface.Condition %in% c("Mud, Dirt, Gravel", "MUD, DIRT, GRAVEL") ~ "Mud/Dirt",
    Surface.Condition %in% c("Sand", "SAND", "OIL", "Other", "OTHER") ~ "Other",
    TRUE ~ "Unknown"
  )) %>%
  count(Collision.Type, Simplified_Surface) %>%
  group_by(Collision.Type) %>%
  mutate(percentage = n / sum(n))

ggplot(simplified_barplot_data, aes(x = Collision.Type, y = percentage, fill = Simplified_Surface)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Distribution of Simplified Surface Conditions by Collision Type",
       x = "Collision Type", y = "Percentage",
       fill = "Surface Condition") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```
There is no definitive way to predict the time of year crashes will occur, but the trend shows a notable spike in crashes from January to March in 2015 and 2017. Additionally, there is often an increase in crashes in the month of November.

```{r p3}
# Monthly trend of Crash COunts by Year

library(dplyr)
library(ggplot2)
library(lubridate)


crash_data <- cleaned_dataset %>%
  mutate(
    Crash_Date = ymd_hms(Crash.Date.Time), 
    Year = year(Crash_Date),
    Month = month(Crash_Date, label = TRUE)
  ) %>%
  filter(!is.na(Crash_Date))  

monthly_data <- crash_data %>%
  group_by(Year, Month) %>%
  summarise(Crash_Count = n(), .groups = 'drop')

ggplot(monthly_data, aes(x = Month, y = Crash_Count, group = Year)) +
  geom_line() +
  labs(
    title = "Monthly Trend of Crash Counts by Year",
    x = "Month", y = "Number of Crashes"
  ) +
  facet_wrap(~ Year, scales = "free_y") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Out of the top 10 roads with the most crashes the surface conditions are mostly dry. 
```{r p4}
# Find the top 10 roads with the most crashes
top_road_crash_data <- cleaned_dataset %>%
  filter(!is.na(`Road.Name`)) %>%
  count(`Road.Name`, sort = TRUE) %>%
  top_n(10, n)

# Filter the cleaned dataset to only include these top 10 roads
top_road_conditions <- cleaned_dataset %>%
  filter(`Road.Name` %in% top_road_crash_data$`Road.Name`) %>%
  mutate(Simplified_Surface = case_when(
    Surface.Condition %in% c("Dry", "DRY") ~ "Dry",
    Surface.Condition %in% c("Wet", "WET", "Water (standing, moving)", "WATER(STANDING/MOVING)") ~ "Wet",
    Surface.Condition %in% c("Ice", "ICE", "Ice/Frost", "Snow", "SNOW", "Slush", "SLUSH") ~ "Snow/Ice",
    Surface.Condition %in% c("Mud, Dirt, Gravel", "MUD, DIRT, GRAVEL") ~ "Mud/Dirt",
    Surface.Condition %in% c("Sand", "SAND", "OIL", "Other", "OTHER") ~ "Other",
    TRUE ~ "Unknown"
  )) %>%
  count(`Road.Name`, Simplified_Surface) %>%
  group_by(`Road.Name`) %>%
  mutate(percentage = n / sum(n))

# Plot the distribution of surface conditions for the top 10 roads
ggplot(top_road_conditions, aes(x = reorder(`Road.Name`, -n), y = percentage, fill = Simplified_Surface)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  coord_flip() +
  labs(
    title = "Surface Conditions for the Top 10 Roads with the Most Crashes",
    x = "Road Name", y = "Percentage of Crashes",
    fill = "Surface Condition"
  ) +
  theme_minimal()
```

